---
title: "getting started"
output: 
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{getting_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# `fellingdateR`: Estimate, report and combine felling dates of historical tree-ring series

This R-package offers a set of functions that can assist you to infer felling date estimates from dated tree-ring series. The presence of (partially) preserved sapwood or waney edge allows to estimate a range for the actual felling date, for individual series as well as for a group of timbers. Furthermore, an additional function provides a tool to sum sapwood probability distributions, comparable to 'summed probability densities' commonly applied to sets of radiocarbon (^14^C) dates.

Where it can be assumed that a group of historical timbers were all felled at the same time (i.e. in the same year), but due to the absence of the bark/cambial zone (waney edge) and the last formed tree ring this cannot be assessed, the preserved sapwood rings can be used to infer a date range for the felling date.

Taking into account the observed number of sapwood rings on all samples and combining them into a single estimate, is likely to provide a more accurate and precise estimate of the felling date year for the group of timbers under study.

This vignette provides a quick overview of the package main functions.

## Installation

You can install the development version of fellingdateR from [GitHub](https://github.com/) with:

``` {r eval=FALSE}
#install.packages("devtools")
devtools::install_github("hanecakr/fellingdateR")
```
or

```{r eval=FALSE}
# install.packages("pak")
pak::pak("hanecakr/fellingdateR")
```


## Basic example

In the following example the combined felling date range for a set of five dated tree-ring series is computed:

```{r setup}
library(fellingdateR)

```

```{r example}
## a data set where all series have partially preserved sapwood:
sw_combine(trs_example1, plot = TRUE)

```

The light grey distributions represent the probability density function of the felling date range for each individual series. The dark grey distribution is the combined estimate for a common felling date. 

The sapwood data used in the example above to estimate the felling date range, was published by Hollstein in 1980:

```{r model_sapwood_counts}
sw_model("Hollstein_1980")
```


## Main functions

### sw_interval

This function computes the probability density function (PDF) and highest probability density interval (hdi) of the felling date range based on the observed number of sapwood rings, their chronological dating and the selected sapwood data and model.

In the example below, 10 sapwood rings were observed on a sample (last ring dated to 1234 AD) that is supposed to have a provenance in the Southern Baltic region (sapwood model published by Wazny, 1990). The hdi delineates an interval in which the actual felling date is most likely situated. It is the shortest interval within a probability distribution for a given probability mass or credible interval. The hdi summarizes the distribution by specifying an interval that spans most of the distribution, say 95% of it, as such that every point inside the interval has higher credibility than any point outside the interval.

```{r interval-individual}
sw_interval(n_sapwood = 10,
            last = 1234,
            hdi = TRUE,
            credMass = .95, 
            sw_data = "Wazny_1990", 
            densfun = "lognormal")
```

When `hdi = FALSE` a matrix is returned with scaled p values for each number of observed sapwood rings.

The results of this procedure can be visualized by setting `plot = TRUE`.

```{r }
# 10 sapwood rings observed and the Wazny 1990 sapwood model:
sw_interval(n_sapwood = 10,
            last = 1234,
            hdi = TRUE,
            credMass = .95, 
            sw_data = "Wazny_1990", 
            densfun = "lognormal",
            plot = T)
```

### fd_report

Reporting estimates of the felling date range for individual series, is provided by the `fd_report()` function.

```{r}
tmp <- data.frame(id = c("aaa", "bbb", "ccc"),
                 swr = c(10, 11, 12),
                 waneyedge = c(FALSE, FALSE,TRUE),
                 end = c(123, 456, 1789))
fd_report(tmp,
         series = "id",
         n_sapwood = "swr",
         last = "end",
         sw_data = "Wazny_1990")
```

### sw_combine

The procedure to combine felling dates of a group of related, individual series with (partially) preserved sapwood, in order to narrow down the range of a common felling date is provided by the `sw_combine()` function. It returns a list with:

-   the pdf for the felling date of the individual series and the pdf of the model that combines these individual series (`$dataRaw`),

-   the hdi for the combined estimate of the common felling date (`$hdi_model`),

-   the *Agreement index* (`$A_comb`) of the model, expressing how well the individual series fit into the model (ideally around 100%, and not lower than the critical threshold A_c = 60%) ,

-   an overview of the felling date range for the individual series (`$individual_series`), and their *Agreement index* (A_i) to the combined model.


#### trs_example0

A data set with dated tree-ring series which all have partially preserved sapwood. The names of the variables in the data set are mapped to the parameters of the `sw_combine()` function. In the example below, the numeric output is returned:

```{r}

# In trs_example0, the column names are not equivalent to the default names in `sw_combine()`.

output_comb <- sw_combine(trs_example0,
   series = "trs",
   last = "end",
   n_sapwood = "swr",
   waneyedge = "bark",
   credMass = .954,
   plot = FALSE
)

head(output_comb$rawData, 20)

output_comb[-1]
```



#### trs_example2

A data set with 5 tree-ring series of which one has an exact felling date:

```{r}
sw_combine(trs_example2, plot = TRUE)

```

#### trs_example3

A data set where all tree-ring series have been measured up to the waney edge:  

```{r }
sw_combine(trs_example3, plot = TRUE)
```

#### trs_example4

An attempt to compute a common felling date for a group of tree-ring series. All series include partially preserved sapwood:

```{r }
sw_combine(trs_example4, plot = TRUE)
```
No common felling date can be computed for this particular data set. The model fails (agreement index of the model is  below the critical threshold value (A_c) of 60%). Three series have an individual agreement index below 60%.

#### trs_example5

When no sapwood rings are observed and measured, only an `earliest possible felling date` (*terminus post quem*) can be determined: 

```{r }
sw_combine(trs_example5, plot = TRUE)
```

### sw_sum

The `sw_sum()` function computes the the summed probability density (SPD) for a set of felling date ranges.

```{r }
sw_sum(trs_example7, plot = TRUE)
```



## Helper functions

### sw_data_overview

The function `sw_data_overview()` provides an overview of all published sapwood data sets that are distributed with this package.

```{r overview}
sw_data_overview()

```

Load the original data with, e.g. `get("Hollstein_1980")`.

### sw_data_info

More information on one of the sapwood data sets - how to cite the data set, the area the data represents, the number of observations and some basic summary stats - can be retrieved by the `sw_data_info()` function.

```{r info}
sw_data_info("Pilcher_1987")

```

### sw_model

A graphical representation of a sapwood data set is provided by the `sw_model()` function.

```{r model}

sw_model("Sohar_2012_ELL_c")

sw_data_info("Sohar_2012_ELL_c")

```

The `sw_model()` function allows to fit a distribution to a data set of observed sapwood numbers and computes the highest posterior density interval (hdi) for a given credibility mass. The density function fitted to the sapwood data set should be one of:

-   "lognormal" (the default value),
-   "normal",
-   "weibull",
-   "gamma".

The credible interval should be a value between 0 and 1.

```{r sw_model_plot2}
sw_model("Wazny_1990", densfun = "gamma", credMass= .90, plot = TRUE)

```

When `plot = FALSE`, a list with the numeric output of the modelling process is returned.



### read_fh and get_header

The `read_fh()` function is an extension to the `dplR::read.fh()` function from the [**dplR package**](https://github.com/AndyBunn/dplR) ([Bunn 2008](https://doi.org/10.1016/j.dendro.2008.01.002), [Bunn 2010](https://doi.org/10.1016/j.dendro.2009.12.001), [Bunn *et al.* 2022](https://github.com/AndyBunn/dplR)). It allows to read .fh (format Heidelberg) files of ring widths AND additional information found in the HEADER fields are listed as attributes.

This function also works with ring-width data in CHRONO and HALF-CHRONO format.

Furthermore, the `read_fh()` function is case insensitive.


In the example below, an .fh file with dated tree-ring series from a medieval ship DOEL1 is read with `read_fh()`. The ring width measurement, in different formats, are converted to a `data.frame`. 

```{r}
Doel1 <- system.file("extdata", "DOEL1.fh", package = "fellingdateR")

Doel1_trs <- read_fh(Doel1, verbose = FALSE)

head(Doel1_trs, 15)

```

When `header = TRUE`, the `get_header()` function is triggered and HEADER fields in the .fh file are returned as a `data.frame`, instead of the ring-width measurements.

```{r}
read_fh(Doel1, verbose = FALSE, header = TRUE)

```

The `data.frame` with the HEADER fields can then be used as input for the sw_functions:

```{r fig.asp=1}
Doel1_header <- read_fh(Doel1, verbose = FALSE, header = TRUE)

Doel1_header |> 
   # first convert column 'waneyedge' to a logical vector 
   # (in the original .fh file header fields the presence of waney edge is
   # indicated by "WK" - a character string - in the corresponding HEADER field)
   dplyr::mutate(waneyedge = dplyr::if_else(grepl("wk", waneyedge, ignore.case = TRUE),
                               TRUE, FALSE)
   ) |> 
   sw_combine(plot = TRUE)
   
```

The attempt to combine the dated tree-ring series from DOEL1 into a single felling-date range fails. These tree-ring series cannot represent a single event.

However, leaving out the tree-ring series of the keelplank (series "K1_091") of this medieval ship shows that the hull planking and some frame elements could share a common felling date:

```{r fig.asp=1}
Doel1_header |> 
   dplyr::mutate(waneyedge = dplyr::if_else(grepl("wk", waneyedge, ignore.case = TRUE),
                               TRUE, FALSE)
   ) |> 
   dplyr::filter(series != "K1_091") |> 
   sw_combine(plot = TRUE)
```

For more details see: Haneca K. & Daly A. (2014). Tree-Rings, Timbers and Trees: a dendrochronological survey of the 14th-century cog, Doel 1. _International Journal of Nautical Archaeology_ 43, 87–102. https://doi.org/10.1111/1095-9270.12037


### cor_table

The `cor_table()` function computes common correlation values between dated tree-ring series (x) and a set of reference chronologies (y). This function assists in checking the end date of the series against absolutely dated reference chronologies, but also to select the most appropriate sapwood model for your data according to the provenance of the wood (by running it against chronologies that represent a geographically confined region).

The correlation values computed are:

-   glk: 'Gleichläufigkeit' or 'percentage of parallel variation' (Buras & Wilmking 2015; Eckstein & Bauch 1969; Huber 1942; Visser 2020)

-   glk_p: significance level associated with the glk-value (Jansma 1995)

-   r_pearson: the Pearson's correlation coefficient

-   t_St: Student's t-value

-   t_BP: t-values according to the Baillie & Pilcher (1973) algorithm

-   t_Ho: t-values according to the Hollstein (1980) algorithm


```{r}
Doel1 <- system.file("extdata", "DOEL1.fh", package = "fellingdateR")

Doel1_trs <- read_fh(Doel1, verbose = FALSE)

# crossdating ring-width series from Doel 1 against each other:
cor_results <- cor_table(Doel1_trs, output = "table", min_overlap = 80)
head(cor_results, 20)

```

